---
title: 'lme4: Mixed-effects models in R'
author: "Joshua Cook"
date: "10/2/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(lme4)
library(tidyverse)

sleepstudy <- as_tibble(sleepstudy)
```

## vignette: [Fitting Linear Mixed-Effects Models Using lme4](https://cran.r-project.org/web/packages/lme4/vignettes/lmer.pdf)

### Data

The vignette used a data set on the average reactoin time per day for subjects in a sleep deprevation study.

```{r}
sleepstudy
```

On day 0, the subjects had a normal amount of sleep, but then were restricted to 3 hours after that.
`Reaction` represents the average reaction time (in ms) on a series of tests.


```{r}
ggplot(sleepstudy, aes(x = Days, y = Reaction)) +
    geom_smooth(method = "lm", alpha = 0.7, color = "grey40") +
    geom_point(aes(color = Reaction)) +
    facet_wrap(~Subject, nrow = 3) +
    scale_color_viridis_c() +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(y = "Reaction time (ms)",
         x = "Days of sleep deprivation (3 hrs/night)",
         title = "Overview of `sleepstudy` data")
```

### Model

The subject was treated as a random effect in the following model.

```{r}
fm1 <- lmer(Reaction ~ Days + (Days | Subject),
            sleepstudy)
summary(fm1)
```

The coefficients of the fixed effects intercept and slope are 251 ms and 10 ms.
These are estimating the population mean values of the random intercept and slope.
