---
title: 'lavaan: latent variable analysis'
author: "Joshua Cook"
date: "9/30/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(lavaan)
library(tidyverse)
```

['lavaan' Tutorial](http://lavaan.ugent.be/tutorial/index.html)

# A CFA Example

[Confirmatory factor analysis](https://en.wikipedia.org/wiki/Confirmatory_factor_analysis) (CFA) can be done using the `cfa()` function.
CFA is used to test how well a researher's beliefs about how a system works fit the observed data[^1].

[^1]: "Confirmatory factor analysis", *Wikipedia*, 2019-09-30

## The data set

This example will use the provided `HolzingerSwineford1939` data set.
It contains the 9 mental ability test scores of 7th and 8th graders from two different schools.
Each row is an individual student.

```{r}
as_tibble(HolzingerSwineford1939)
```


The *a priori* belief, in this case,  is that there exists 3 latent variables:

* a **visual** factor is measured by `x1`, `x2`, and `x3`
* a **textual** factor is measured by `x4`, `x5`, and `x6`
* a **spead** factor is measured by `x7`, `x8`, and `x9`

These factors interact with each other, too.

## The 'lavaan' model syntax

The `=~`, read as "*measured by*", is special syntax for defining latent factors. The general format is as follows.

```txt
latent variable =~ indicator1 + indicator2 + indicator3
```
For this example, that translates to the following.

```txt
visual =~ x1 + x2 + x3
textual =~ x4 + x5 + x6
speed =~ x7 + x8 + x9
```

## Fitting the model

For the function, the model *must be defined within **single** quotes.*

```{r}
HS.model <- '
    visual =~ x1 + x2 + x3
    textual =~ x4 + x5 + x6
    speed =~ x7 + x8 + x9
'
```

Now the model is fit using `cfa()`.

```{r}
fit <- cfa(HS.model, data = HolzingerSwineford1939)
```


## The results

```{r}
summary(fit, fit.measures = TRUE)
```

The header has the following information:

* the 'lavaan' version number
* did 'lavaan' converge normally or not, and how many iterations were needed
* the number of observations that were effectively used in the analysis
* the estimator that was used to obtain the parameter values (here: `ML`)
* the model test statistic, the degrees of freedom, and a corresponding p-value

Various paramters of the model, including a p-value, are displayed in the following sections.

The "Paramter Estimates" section shows the various free and fixed paramters in the model. 




# A SEM Example



## The data set

This example uses the built in `PoliticalDemocracy` data set.
The columns show various indicators of freedom in different years.
(For more information see the help page `?PoliticalDemocracy`.)

```{r}
as_tibble(PoliticalDemocracy)
```

## The model

The model we want to create is as follows:

- `y1` through `y4` link to the latent factor `dem60`
- `y5` through `y8` link to the latent factor `dem65`
- `x1` through `x3` link to the latent factor `ind60`
- `dem60` is linked to `ind60`
- `dem65` is link to `dem60` and `ind60`

In `lavaan` syntax, this translates to the following.

```{r}
model <- '
    # measurement model
        ind60 =~ x1 + x2 + x3
        dem60 =~ y1 + y2 + y3 + y4
        dem65 =~ y5 + y6 + y7 + y8
    # regressions
        dem60 ~ ind60
        dem65 ~ ind60 + dem60
    # residual correlations
        y1 ~~ y5
        y2 ~~ y4 + y6
        y3 ~~ y7
        y4 ~~ y8
        y6 ~~ y8
'
```

In the above model, the latent variables are defined using `=~`, regressions between the latent variables are defined using `~`, and covariance formulas are expressed using the `~~` operator.
The last case, where two variables are connected using the `~~` operator, indicates that the variables can be correlated.
For example, `y1 ~~ y5` is included becasue the "Expert ratings of the freedom of the press in 1960" may be correlated with "Expert ratings of the freedom of the press in 1965."
According to the tutorial, "[t]his is sometimes done if it is believed that the two variables have something in common that is not captured by the latent variables."
In addition, the expression `y2 ~~ y4 + y6` is just shorthand for `y2 ~~ y4` and `y2 ~~ y6`.


## Fitting the model

```{r}
fit <- sem(model, data = PoliticalDemocracy)
summary(fit, standardized = TRUE)
```

To demonstrate that the model fitting is sensitive to randomly added latent variables, I add a new one called `myLV` and defined it with 4 random variables.

```{r}
my_model <- '
    # measurement model
        ind60 =~ x1 + x2 + x3
        dem60 =~ y1 + y2 + y3 + y4
        dem65 =~ y5 + y6 + y7 + y8
        myLV =~ x1 + y1 + y3 + y8
    # regressions
        dem60 ~ ind60
        dem65 ~ ind60 + dem60
    # residual correlations
        y1 ~~ y5
        y2 ~~ y4 + y6
        y3 ~~ y7
        y4 ~~ y8
        y6 ~~ y8
'
summary(sem(my_model, data = PoliticalDemocracy), standardized = TRUE)
```

From the summary, it is obvious that there is no latent variable `myLV` as currently defined.


# Additional Syntax

## Fixing parameters

To fix a parameter in 'lavaan', multiply it by a numeric value.
`y2` is fixed in the following model.

```txt
f =~ y1 + 1*y2 + y3
```

Can specify that the covariance of latent factors are 0 by pre-multiplying by 0.
The following model sets the covariance between the latent variables `visual` and `textual` with `speed` to zero.
The variance of `speed` is also fixed.
The variances of `visual` and `textual` are left free by default.

```txt
# three-factor model
    visual =~ x1 + x2 + x3
    textual =~ x4 + x5 + x6
    speed   =~ NA*x7 + x8 + x9
# orthogonal factors
    visual ~~ 0*speed
    textual ~~ 0*speed
# fix variance of speed factor
    speed ~~ 1*speed
```

## Simple equality constraints

It is possible to set equality constraints on free parameters.
This is useful if we have an * priori* reason to believe that the indicators are equivalent.
It can be daccomplished one by assigning the parameters the same label (a label is assigned by multiplying the variable against a character string).
In the following example, `x2` and `x3` are set to have the same parameter by assigning them both the label `"v1"`.

```txt
 visual =~ x1 + v2*x2 + v2*x3
textual =~ x4 + x5 + x6
  speed =~ x7 + x8 + x9
```

## Nonlinear equality and inequality constraints

The following demonstration will use this example regression model.
The regression coefficients have been explicitly named `"b1"`, `"b2"`, and `"b3"`.

```txt
y ~ b1*x1 + b2*x2 + b3*x3
```

Using a toy dataset.

```{r}
set.seed(1234)
Data <- data.frame(y = rnorm(100), 
                   x1 = rnorm(100), 
                   x2 = rnorm(100),
                   x3 = rnorm(100))
model <- ' y ~ b1*x1 + b2*x2 + b3*x3 '
fit <- sem(model, data=Data)
coef(fit)
```

Suppose we wanted to set the following conrtaints on the parameters:

$$
b_1: b_1 = (b_2+b_3)^2 \quad \text{and} \quad b_1 \geq \exp(b_2 + b_3)
$$

These can be explicitly expressed as follows.

```{r}
model.constr <- '
# model with labeled parameters
    y ~ b1*x1 + b2*x2 + b3*x3
# constraints
    b1 == (b2 + b3)^2
    b1 > exp(b2 + b3)
'
```

Refitting the model shows that the constraints were effective.

```{r}
fit <- sem(model.constr, data = Data)
coef(fit)

# inspect that the constraints were observed
a <- coef(fit)
(a[["b2"]] + a[["b3"]])^2
exp(a[["b2"]] + a[["b3"]])
```
